Memory for television episodes preserves event content while introducing new across-event similarities

### 2020-02-07 paper/main.tex
progress on methods

\subsection*{Modeling the dynamic content of the video and recall transcripts}
\subsubsection*{Topic modeling}
The input to the topic model we trained to characterize the dynamic content of the video comprised 998 hand-generated annotations of short (mean: 2.96s) scenes spanning the video clip~(\citealp{ChenEtal17} generated 1000 annotations total; we removed two referring to the break between the first and second scan sessions, during which no fMRI data was collected).  We concatenated the text for all of the annotated features within each segment, creating a ``bag of words'' describing each scene and performed some minor preprocessing (e.g., stemming possessive nouns and removing punctuation).  We then re-organized the text descriptions into overlapping sliding windows spanning (up to) 50 scenes each.  In other words, we created a ``context" for each scene comprising the text descriptions of the preceding 25 scenes, the present scene, and the following 24 scenes.  To model the ``context" for scenes near the beginning and end of the video (i.e., within 25 scenes of the beginning or end), we created overlapping sliding windows that grew in size from one scene to the full length, then similarly tapered their length at the end.  This additionally ensured that each scene's content was represented in the text corpus an equal number of times.
% In other words, one sample contained the text from scenes 1-50, the next contained the text from scenes 2-51, and so on.  In order to ensure each scene's content was represented in these 
% ------ NOTE: ------
% need to refine explanation here a little bit
% - Paxton

@@

We used an optimization procedure to select the appropriate $K$ for each topic proportions matrix.  Prior studies on narrative structure and processing have shown that we both perceive and internally represent the world around us at multiple, hierarchical timescales (e.g., \citealp{HassEtal08, LernEtal11, HassEtal15, ChenEtal17, BaldEtal17, BaldEtal18}).  However, for the purposes of our framework, we sought to identify the single timeseries of event-representations that is emphasized \textit{most heavily} in the temporal structure of the video and of each participant's recall.  We quantified this as the set of $K$ states that maximized the similarity between topic vectors for timepoints comprising each state, while minimizing the similarity between topic vectors for timepoints across different states.  Specifically, we computed (for each matrix)
\[
  \argmax_K \left[W_{1}(a, b)\right],
\]
where $a$ was the distribution of within-state topic vector correlations, and $b$ was the distribution of across-state topic vector correlations .  We computed the first Wasserstein distance ($W_{1}$; also known as ``earth mover's distance"; \citealp{Dobr70, RamdEtal17}) between these distributions for a large range of possible $K$-values (range [2,50]), and selected the $K$ that yielded the greatest difference.  Figure~\ref{fig:model}B displays the event boundaries returned for the video, and Figure~\corrmats~displays the event boundaries returned for each participant's recalls (See Fig.~\kopt~for the optimization functions for the video and recalls).  After obtaining these event boundaries, we created stable estimates of each topic proportions matrix by averaging the topic vectors within each event.  This yielded a number-of-events by number-of-topics matrix for the video and recalls from each participant.

@@

Perhaps the simplest and most widely used measure of memory performance is \textit{accuracy}---i.e., the proportion of studied (experienced) items (in this case, the 30 video events) that the participant later remembered.  \cite{ChenEtal17} developed a human rating system whereby the quality of each participant's memory was evaluated by an independent rater.  We found a strong across-participants correlation between these independant ratings and the overall number of events that our HMM approach identified in participants' recalls (Pearson's $r(15) = 0.65, p = 0.004$).

As described below, we next considered a number of memory performance measures that are typically associated with list-learning studies.  We also provide a software package, \texttt{Quail}, for carrying out these analyses~\citep{HeusEtal17b}.

\paragraph{Probability of first recall (PFR).}  PFR curves~\citep{WelcBurn24, PostPhil65, AtkiShif68} reflect the probability that an item will be recalled first as a function of its serial position during encoding. To carry out this analysis, we initialized a number-of-participants (17) by number-of-video-events (30) matrix of zeros. Then for each participant, we found the index of the video event that was recalled first (i.e., the video event whose topic vector was most strongly correlated with that of the first recall event) and filled in that index in the matrix with a 1.  Finally, we averaged over the rows of the matrix, resulting in a 1 by 30 array representing the proportion of participants that recalled an event first, as a function of the order of the event's appearance in the video (Fig.~\ref{fig:list-learning}A).

\paragraph{Lag conditional probability curve (lag-CRP).} The lag-CRP curve~\citep{Kaha96} reflects the probability of recalling a given event after the just-recalled event, as a function of their relative positions (or \textit{lag}).  In other words, a lag of 1 indicates that a recalled event came immediately after the previously recalled event in the video, and a lag of -3 indicates that a recalled event came 3 events before the previously recalled event.  For each recall transition (following the first recall), we computed the lag between the current recall event and the next recall event, normalizing by the total number of possible transitions.  This yielded a number-of-participants (17) by number-of-lags (-29 to +29; 61 lags total) matrix. We averaged over the rows of this matrix to obtain a group-averaged lag-CRP curve (Fig.~\ref{fig:list-learning}B).

\paragraph{Serial position curve (SPC).} SPCs~\citep{Murd62a} reflect the proportion of participants that remember each item as a function of the items' serial position during encoding. We initialized a number-of-participants (17) by number-of-video-events (30) matrix of zeros. Then, for each recalled event, for each participant, we found the index of the video event that the recalled event most closely matched (via the correlation between the events' topic vectors) and entered a 1 into that position in the matrix (i.e., for the given participant and event). This resulted in a matrix whose entries indicated whether or not each event was recalled by each participant (depending on whether the corresponding entires were set to one or zero).  Finally, we averaged over the rows of the matrix to yield a 1 by 30 array representing the proportion of participants that recalled each event as a function of the order of the event's appearance in the video (Fig.~\ref{fig:list-learning}C).

\paragraph{Temporal clustering scores.} Temporal clustering describes participants' tendency to organize their recall sequences by the learned items' encoding positions.  For instance, if a participant recalled the video events in the exact order they occurred (or in exact reverse order), this would yield a score of 1.  If a participant recalled the events in random order, this would yield an expected score of 0.5.  For each recall event transition (and separately for each participant), we sorted all not-yet-recalled events according to their absolute lag (i.e., distance away in the video).  We then computed the percentile rank of the next event the participant recalled.  We averaged these percentile ranks across all of the participant's recalls to obtain a single temporal clustering score for the participant.

\paragraph{Semantic clustering scores.} Semantic clustering describes participants' tendency to recall semantically similar presented items together in their recall sequences.  Here, we used the topic vectors for each event as a proxy for its semantic content. Thus, the similarity between the semantic content for two events can be computed by correlating their respective topic vectors.  For each recall event transition, we sorted all not-yet-recalled events according to how correlated the topic vector \textit{of the closest-matching video event} was to the topic vector of the closest-matching video event to the just-recalled event.  We then computed the percentile rank of the observed next recall.  We averaged these percentile ranks across all of the participant's recalls to obtain a single semantic clustering score for the participant.

% I moved the actual results from the two paragraphs to the results section as an afterthought to the list-learning analyses.  We could also possibly just take them out, though we should keep them if you think it's worth referencing as support for our argument (see my note in discussion section)
% -Paxton

% NOTE: Thinking about removing this analysis? More of a sanity check than something novel and cool, and takes up space. Happy to be convinced otherwise - andy
% leaving this one to Jeremy's judgement
% -Paxton

% To quantify the similarity between the video topic trajectory and individual recall topic trajectories, we considered several novel metrics.  First, we tested whether each participant's recall trajectory matched the video trajectory in a general sense. For each participant, we filtered the video trajectory to only include the events that the participant remembered.  We then computed the root mean squared difference (RMSD) between the remaining video events and the (closest-matching) recalled events.  For example, if the topic vectors for a participant's recall event topic vectors matched the corresponding video event topic vectors exactly (and in order), the expected RMSD for those events would be 0.  However, if the participant's recall events did not perfectly match the video events, or if they were out of order, then the RMSD would be greater than 0.  To assess the significance of the match between the video and recall trajectories, we carried out a permutation procedure whereby, for each of 10000 repetitions, we circularly shifted the recall trajectories (in time) by a random amount and then re-computed the RMSD each time.  This yielded a distribution of ``null'' RMSD values for each participant.  The observed RMSD values reached significance (i.e., $p < 0.05$, reflecting that more than 95\% of the null RMSD values were greater than the observed RMSD value) for nine of the participants (3, 4, 8--13, and 17).  (For the remaining participants this test yielded $0.05 \leq p < 0.25$.)  The observed RMSD values were also reliably correlated with hand-annotated memory performance across participants (Pearson's $r(15) = -0.57, p = 0.016$).  In other words, a closer match between the video and recall topic trajectories was related to better overall recall performance.

@@

\paragraph*{Precision.}
We tested whether participants who recalled more events were also more \textit{precise} in their recollections. For each participant, we computed the average correlation between the topic vectors for each recall event and those of its closest-matching video event. This gave a single value per participant representing the average precision across all recalled events.  We then Fisher's $z$-transformed these values and correlated them with both hand-annotated and model-derived (i.e., $k$ or the number of events recovered by the HMM) memory performance.

\paragraph*{Distinctiveness.}
We also considered the \textit{distinctiveness} of each recalled event. That is, how uniquely a recalled event's topic vector matched a given video event topic vector, versus the topic vectors for the other video events. We hypothesized that participants with high memory performance might describe each event in a more distinctive way (relative to those with lower memory performance who might describe events in a more general way).  To test this hypothesis we define a distinctiveness score for each recall event as

\[
  d(\mathrm{event}) = 1 - \bar{c}(\mathrm{event}),
\]
where $\bar{c}(\mathrm{event})$ is the average correlation between the given recalled event's topic vector and the topic vectors from all video events \textit{except} the best-matching video event.  We then averaged these distinctiveness scores across all of the events recalled by the given participant. As above, we used Fisher's $z$-transformation before correlating these values with hand-annotated and model derived memory performance scores across-subjects.

>>>>

\subsection*{Modeling the dynamic content of the video and recall transcripts}
\subsubsection*{Topic modeling}
The input to the topic model we trained to characterize the dynamic content of the video comprised 998 hand-generated annotations of short (mean: 2.96s) scenes spanning the video clip~(\citealp{ChenEtal17} generated 1000 annotations total; we removed two referring to the break between the first and second scan sessions, during which no fMRI data was collected).  We concatenated the text for all of the annotated features within each segment, creating a ``bag of words'' describing each scene and performed some minor preprocessing (e.g., stemming possessive nouns and removing punctuation).  We then re-organized the text descriptions into overlapping sliding windows spanning (up to) 50 scenes each.  In other words, we created a ``context" for each scene comprising the text descriptions of the preceding 25 scenes, the present scene, and the following 24 scenes.  To model the ``context" for scenes near the beginning and end of the video (i.e., within 25 scenes of the beginning or end), we created overlapping sliding windows that grew in size from one scene to the full length, then similarly tapered their length at the end.  This additionally ensured that each scene's content was represented in the text corpus an equal number of times.
% ------ NOTE: ------
% need to refine explanation here a little bit
% - Paxton

@@

We used an optimization procedure to select the appropriate $K$ for each topic proportions matrix.  Prior studies on narrative structure and processing have shown that we both perceive and internally represent the world around us at multiple, hierarchical timescales (e.g., \citealp{HassEtal08, LernEtal11, HassEtal15, ChenEtal17, BaldEtal17, BaldEtal18}).  However, for the purposes of our framework, we sought to identify the single timeseries of event-representations that is emphasized \textit{most heavily} in the temporal structure of the video and of each participant's recall.  We quantified this as the set of $K$ states that maximized the similarity between topic vectors for timepoints comprising each state, while minimizing the similarity between topic vectors for timepoints across different states.  Specifically, we computed (for each matrix)
\[
  \argmax_K \left[W_{1}(a, b)\right],
\]
where $a$ was the distribution of within-state topic vector correlations, and $b$ was the distribution of across-state topic vector correlations .  We computed the first Wasserstein distance ($W_{1}$; also known as ``earth mover's distance"; \citealp{Dobr70, RamdEtal17}) between these distributions for a large range of possible $K$-values (range [2,50]), and selected the $K$ that yielded the maximum value.  Figure~\ref{fig:model}B displays the event boundaries returned for the video, and Figure~\corrmats~displays the event boundaries returned for each participant's recalls.  See Figure \kopt~for the optimization functions for the video and recalls.  After obtaining these event boundaries, we created stable estimates of the content represented in each event by averaging the topic vectors across timepoints between each pair of event boundaries.  This yielded a number-of-events by number-of-topics matrix for the video and recalls from each participant.

@@

Perhaps the simplest and most widely used measure of memory performance is \textit{accuracy}---i.e., the proportion of studied (experienced) items (in this case, video events) that the participant later remembered.  \cite{ChenEtal17} used this method to rate each participant's memory quality by computing the proportion of (50, manually identified) scenes mentioned in their recall.  We found a strong across-participants correlation between these independent ratings and the proportion of (30, HMM-identified) video events matched to participants' recalls (Pearson's $r(15) = 0.71, p = 0.002$).  We further considered a number of more nuanced memory performance measures that are typically associated with list-learning studies.  We also provide a software package, \texttt{Quail}, for carrying out these analyses~\citep{HeusEtal17b}.

\paragraph{Probability of first recall (PFR).}  PFR curves~\citep{WelcBurn24, PostPhil65, AtkiShif68} reflect the probability that an item will be recalled first as a function of its serial position during encoding. To carry out this analysis, we initialized a number-of-participants (17) by number-of-video-events (30) matrix of zeros. Then for each participant, we found the index of the video event that was recalled first (i.e., the video event whose topic vector was most strongly correlated with that of the first recall event) and filled in that index in the matrix with a 1.  Finally, we averaged over the rows of the matrix, resulting in a 1 by 30 array representing the proportion of participants that recalled an event first, as a function of the order of the event's appearance in the video (Fig.~\ref{fig:list-learning}A).
% ------ NOTE: ------
% reiterate meaning of error ribbons in list-learning figure? (already noted in figure caption)
% - Paxton

\paragraph{Lag conditional probability curve (lag-CRP).} The lag-CRP curve~\citep{Kaha96} reflects the probability of recalling a given item after the just-recalled item, as a function of their relative encoding positions (or \textit{lag}).  In other words, a lag of 1 indicates that a recalled item was presented immediately after the previously recalled item, and a lag of -3 indicates that a recalled item came 3 items before the previously recalled item.  For each recall transition (following the first recall), we computed the lag between the current recall event and the next recall event, normalizing by the total number of possible transitions.  This yielded a number-of-participants (17) by number-of-lags (-29 to +29; 61 lags total) matrix. We averaged over the rows of this matrix to obtain a group-averaged lag-CRP curve (Fig.~\ref{fig:list-learning}B).

\paragraph{Serial position curve (SPC).} SPCs~\citep{Murd62a} reflect the proportion of participants that remember each item as a function of the items' serial positions during encoding. We initialized a number-of-participants (17) by number-of-video-events (30) matrix of zeros. Then, for each recalled event, for each participant, we found the index of the video event that the recalled event most closely matched (via the correlation between the events' topic vectors) and entered a 1 into that position in the matrix. This resulted in a matrix whose entries indicated whether or not each event was recalled by each participant (depending on whether the corresponding entires were set to one or zero).  Finally, we averaged over the rows of the matrix to yield a 1 by 30 array representing the proportion of participants that recalled each event as a function of the events' order appearance in the video (Fig.~\ref{fig:list-learning}C).

\paragraph{Temporal clustering scores.} Temporal clustering describes a participant's tendency to organize their recall sequences by the learned items' encoding positions.  For instance, if a participant recalled the video events in the exact order they occurred (or in exact reverse order), this would yield a score of 1.  If a participant recalled the events in random order, this would yield an expected score of 0.5.  For each recall event transition (and separately for each participant), we sorted all not-yet-recalled events according to their absolute lag (i.e., distance away in the video).  We then computed the percentile rank of the next event the participant recalled.  We averaged these percentile ranks across all of the participant's recalls to obtain a single temporal clustering score for the participant.

\paragraph{Semantic clustering scores.} Semantic clustering describes a participant's tendency to recall semantically similar presented items together in their recall sequences.  Here, we used the topic vectors for each event as a proxy for its semantic content. Thus, the similarity between the semantic content for two events can be computed by correlating their respective topic vectors.  For each recall event transition, we sorted all not-yet-recalled events according to how correlated the topic vector \textit{of the closest-matching video event} was to the topic vector of the closest-matching video event to the just-recalled event.  We then computed the percentile rank of the observed next recall.  We averaged these percentile ranks across all of the participant's recalls to obtain a single semantic clustering score for the participant.

% To quantify the similarity between the video topic trajectory and individual recall topic trajectories, we considered several novel metrics.  First, we tested whether each participant's recall trajectory matched the video trajectory in a general sense. For each participant, we filtered the video trajectory to only include the events that the participant remembered.  We then computed the root mean squared difference (RMSD) between the remaining video events and the (closest-matching) recalled events.  For example, if the topic vectors for a participant's recall event topic vectors matched the corresponding video event topic vectors exactly (and in order), the expected RMSD for those events would be 0.  However, if the participant's recall events did not perfectly match the video events, or if they were out of order, then the RMSD would be greater than 0.  To assess the significance of the match between the video and recall trajectories, we carried out a permutation procedure whereby, for each of 10000 repetitions, we circularly shifted the recall trajectories (in time) by a random amount and then re-computed the RMSD each time.  This yielded a distribution of ``null'' RMSD values for each participant.  The observed RMSD values reached significance (i.e., $p < 0.05$, reflecting that more than 95\% of the null RMSD values were greater than the observed RMSD value) for nine of the participants (3, 4, 8--13, and 17).  (For the remaining participants this test yielded $0.05 \leq p < 0.25$.)  The observed RMSD values were also reliably correlated with hand-annotated memory performance across participants (Pearson's $r(15) = -0.57, p = 0.016$).  In other words, a closer match between the video and recall topic trajectories was related to better overall recall performance.
% ------ NOTE: ------
% Andy was thinking about including this in the revision early on.  It wasn't in the preprint version, and with how much else we have in here, I don't think it's worth including anymore.
% - Paxton

@@

\paragraph*{Precision.}
We tested whether participants who recalled more events were also more \textit{precise} in their recollections. For each participant, we computed the average correlation between the topic vectors for each recall event and those of its closest-matching video event. This gave a single value per participant representing the average precision across all recalled events.  We then correlated these values with both hand-annotated and model-derived (i.e., the number of unique video events matched by a participant's recall events) memory performance.

\paragraph*{Distinctiveness.}
We also considered the \textit{distinctiveness} of each recalled event. That is, how unique a participant's description of a video event was, versus their descriptions of other video events.  We hypothesized that participants with high memory performance might describe each event in a more distinctive way (relative to those with lower memory performance who might describe events in a more general way).  To test this hypothesis we define a distinctiveness score for each recall event as

\[
  d(\mathrm{event}) = 1 - \bar{c}(\mathrm{\mathbb{P} \setminus \{event\}}),
\]
where $\bar{c}(\mathrm{\mathbb{P} \setminus \{event\}})$ is the average correlation between the given recall event's topic vector and the topic vectors from all other recall events not matched to the same video event (for a single participant).  We then averaged these distinctiveness scores across all of the events recalled by the given participant and correlated resulting values with hand-annotated and model derived memory performance scores across-subjects, as above.

\noindent Note: in all instances where we performed statistical tests involving precision or distinctiveness scores, we used Fisher's $z$-transformation \citep{Fish25} to stabilize the variance across the distribution of correlation values prior to performing the test.  Similarly, when averaging precision or distinctiveness scores, we $z$-transformed the scores prior to computing the mean, and inverse $z$-transformed the result.
